Race Science in Mainstream Psychology, 1960-2010

### 2025-01-18 paper/paper.qmd
switch quarto format and fix references

bibliography: [references.bib, refs-manual.bib]

format: 
  elsevier-pdf:  # https://github.com/quarto-journals/elsevier
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    journal:
      name: "NA"
      cite-style: "authoryear"
      layout: "onecolumn"
    include-in-header: 
      text: |
        \usepackage{lineno}
        \linenumbers

@@

By the end of the Second World War, a combination of technical developments in anthropology and genetics, along with revelations of Nazi atrocities, had led most members of the scientific community to reject attempts to scientifically justify racial inequalities and colonialism [@BarkanRetreatScientificRacism1992].  But eugenic thinking and scientific racism persisted throughout the second half of the twentieth century, and recent work has found that white supremacists continue to attempt to use scientific research to justify racial violence [@DusterBackdoorEugenics2003; @PanofskyHowWhiteNationalists2021; @PronczukRacistResearcherExposed2022].

@@

We first propose a distinction between scientific racism, race science, and race science discourse.  *Scientific racism* refers to the social practice of purporting to justify racial inequality and colonialism by appealing to the epistemic authority of science.  *Race science* refers to scientific and/or pseudoscientific research, including research products (journal articles, etc.), that can be utilized for scientific racism.  *Race science discourse* refers to a broader category, any treatment of race science as a legitimate area of scientific research.

@@

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisAppropriationBiological2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin, unbalanced quotation marks in Mehler].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].

In the landmark case *Brown v Board of Education of Topeka* (1954), the US Supreme Court banned *de jure* educational segregation.  The Court's decision relied on expert testimony from psychologists and education researchers; but segregationists also put forward their own experts, including Henry Garrett [@WinstonScienceServiceFar1998; @JacksonScienceSegregationRace2005; @SchafferScientificRacismAgain2007].  After *Brown*, Pioneer funded various segregationist efforts across the United States, including lectures on eugenics by Stanford physicist William Shockley [@JacksonScienceSegregationRace2005].  MQ was created to provide a favorable venue for race scientists to publish their views, on the grounds that an "equalitarian dogma" created a censorious "taboo" against their research in mainstream publications [@TuckerFundingScientificRacism2002; @JacksonMythicalTabooRace2020].

@@

Race science has not been heavily studied using an agnotology framework [though see @MillsRacialContract2014 and @KouranyMightScientificIgnorance2020].  But the qualitative historiography suggests that actors such as the Pioneer Fund have deployed a version of the tobacco strategy to promote misleading public perceptions of scientific uncertainty about (or even scientific support for) racial hierarchy. (Indeed, the origins of the "tobacco strategy" are often dated to a 1954 advertising campaign by the tobacco industry, 17 years after Pioneer was founded. Scientific racists and the eugenics movement might have been deploying the "tobacco strategy" for decades before the tobacco industry.)

The Pioneer Fund itself clearly exemplifies the strategies of supporting friendly research; and figures such as Shockley and Jensen exemplify the strategy of recruiting high-profile, distinguished scientists *[Jackson? on funding Shockley]*. Using agent-based models, *[cailin etc.]* have shown that *[st]*.

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "boundary."  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be "incubated," protected from the destructive critical scrutiny they would receive in the mainstream. Boundary venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research.

MQ does not appear to have ever been considered a mainstream journal; it has always existed firmly on the margins. If MQ provided the "safe space" for "incubating" race science, away from mainstream criticism, then boundary venues would be necessary to disseminate race science into the scientific mainstream. We therefore hypothesized (H1) that scientometric and text mining analyses would show race science discourse appearing first and most prominently in MQ, and appearing in mainstream publications only subsequently.

Specifically, among both (other) academics and the general public, the field of behavior genetics is strongly associated with race science [@PanofskyMisbehavingScienceControversy2014]. We therefore hypothesized (H2) that the journal *Behavior Genetics* (BG) would be a major boundary venue for the dissemination of race science discourse.

@@

After document retrieval and text extraction, we pre-processed the text using the `spaCy` NLP (natural language processing) Python library [@SpaCyIndustrialstrengthNatural2018 version 1.9.0] and the R API `spacyr` [@BenoitSpacyrWrapperSpaCy2020 version 1.2.1].  Specifically, we applied regular expressions to remove header/footer copyright notices and hyphenation, used spaCy to annotate and extract noun phrases (eg, "the intelligence test items"), and then cleaned and standardized these phrases (eg, removing the/an/a, coverting all text to lowercase, and replacing all whitespace with underscores: "intelligence_test_items").  We then counted the occurrence of each noun phrase in the document.  The aggregated "document-term matrix" was stored in Parquet format [@VohraApacheParquet2016] for performance reasons, and written and read using the `arrow` package for R [@RichardsonArrowIntegrationApache2023 version 7.0.0].

@@

A key limitation of the current study stems from the inability of text mining techniques to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from overt and direct to a subtle "color-blind" racism [@Bonilla-SilvaRacismRacistsColorBlind2006]. Color-blind racism represents a change in the primary mechanisms by which white privilege is maintained, using non-racialized language in furtherance of the racial status quo both institutionally and individually, and increasingly reliant on the invisibility of socio-cultural mechanisms to preserve inequalities. Color-blind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" [@SaulDogwhistlesPoliticalManipulation2018].  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the color-blind language of "group differences" [@AlderferScienceNonsciencePsychologists2003].  These cases suggest further that color-blindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles [@GillbornSoftlySoftlyGenetics2016; @WillsAreClustersRaces2017].

@@

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic; text version in table @tbl-silge.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal}

@@

In almost all years, most papers in topic 24 were published in mainstream journals rather than MQ.  Jensen's "How Much Can We Boost IQ and Scholastic Achievement?" [@JensenHowMuchCan1969] was published in 1969 in *Harvard Educational Review* (not included in this study), and the time series indicates that, in the early 1970s, there was an increase in articles in topic 24 in both PR and MQ (the only two journals in our corpus that were active at the time).  MQ shows another sharp increase in the late 1980s; using content analysis, Adams and Pilloud found that psychology was the dominant discipline in MQ in the period 1992-2018 [@AdamsMisAppropriationBiological2021].  PID published multiple papers in topic 24 almost immediately after it was founded, with Int showing a more gradual increase between the mid-1970s and mid-1990s.

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct topic of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as venue for late 20th century race science discourse than mainstream psychology journals, especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].

@@

Over the past few decades, many professional organizations in genetics and anthropology have made formal statements rejecting race as a biologically meaningful concept [@anthropology1996aapa; @american2018ashg; @board1998aaa; @committee2020asa; @FuentesAAPAStatementRace2019; @nhgri2023; @rotimi2022ashg; @wynshawboris2020ashg], undermining the core assumption of race science.  The American Psychological Association (APA) --- whose members are predominantly clinical psychologists --- the Federation of Associations in Behavioral & Brain Sciences (FABBS) --- an organization whose purpose is to provide policy recommendations --- and the editors of *Nature Human Behavior* have openly rejected race as biologically meaningful [@american2021apology; @baron2022fabbs; @ScienceMustRespect2022].  But research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as biologically meaningful when they denounce racism [@aps2021statement; @board2020psychonomic].  Psychology as a discipline maintains space for several mythological race science narratives, such as allegations of a taboo against race and intelligence research [@jackson2021mythical], claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science [@roberts_2022], or arguments that holding race science to the same evidentiary standards as other psychological research is a violation of academic freedom [@herbert2023academic].  And historically several race scientists have secured important gatekeeping positions within the scientific community, such as on the editorial boards of *Intelligence* and *Personality and Individual Differences*.

@@

>>>>

bibliography: [references.yaml, refs-manual.bib]

format: 
  aft-pdf:
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    cite-method: citeproc
    include-in-header: 
        text: |
            \usepackage{lineno}
            \linenumbers
    geometry:
        - left = 1.5in
        - right = 1.5in

#  elsevier-pdf:  # https://github.com/quarto-journals/elsevier
#    keep-tex: true
#    link-citations: true
#    mathspec: true
#    fig.env: "figure*"
#    number-sections: true
#    journal:
#      name: "NA"
#      cite-style: "authoryear"
#      layout: "onecolumn"
#    include-in-header: 
#      text: |
#        \usepackage{lineno}
#        \linenumbers
#    cite_method: "citeproc"

@@

By the end of the Second World War, a combination of technical developments in anthropology and genetics, intellectual campaigns by the likes of Franz Boas and W.E.B. Du Bois, and revelations of Nazi atrocities meant that the dominant position within the scientific community was one that rejected attempts to scientifically justify racial inequalities and colonialism [@BarkanRetreatScientificRacism1992; @YudellRaceUnmaskedBiology2014; @JacksonDarwinismDemocracyRace2017].  But eugenic thinking and scientific racism persisted throughout the second half of the twentieth century, and recent work has found that white supremacists continue to attempt to use scientific research to justify racial violence [@DusterBackdoorEugenics2003; @JacksonScienceSegregationRace2005; @PanofskyHowWhiteNationalists2021; @BirdTypologicalThinkingHuman2024; @PronczukRacistResearcherExposed2022].

@@

We first propose a distinction between scientific racism, race science, and race science discourse.  *Scientific racism* refers to the social practice of purporting to justify racial inequality and colonialism by appealing to the epistemic authority of science.  *Race science* refers to scientific and/or pseudoscientific research, including research products (journal articles, etc.), that can be utilized for scientific racism.  *Race science discourse* refers to a broader category, any treatment of race science as a legitimate area of scientific research.  (@sec-operationalizing discusses how this conception of "race science" might be operationalized in practice. @BirdTypologicalThinkingHuman2024 give a concrete example of how a particular figure was transformed into a scientific racist meme.)

@@

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisappropriationBiologicalAnthropology2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin, unbalanced quotation marks in Mehler].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].

In the landmark case *Brown v Board of Education of Topeka* (1954), the US Supreme Court banned *de jure* educational segregation.  The Court's decision relied on expert testimony from psychologists and education researchers; but segregationists also put forward their own experts, including Henry Garrett [@WinstonScienceServiceFar1998; @JacksonScienceSegregationRace2005; @SchafferScientificRacismAgain2007].  After *Brown*, Pioneer funded various segregationist efforts across the United States, including lectures on eugenics by Stanford physicist William Shockley [@TuckerFundingScientificRacism2002; @JacksonScienceSegregationRace2005].  MQ was created to provide a favorable venue for race scientists to publish their views, on the grounds that an "equalitarian dogma" created a censorious "taboo" against their research in mainstream publications [@TuckerFundingScientificRacism2002; @JacksonMythicalTabooRace2020].

@@

Race science has not been heavily studied using an agnotology framework [though see @MillsRacialContract2014 and @KouranyMightScientificIgnorance2020].  The qualitative historiography suggests that actors such as the Pioneer Fund have deployed a version of the tobacco strategy to promote misleading public perceptions of scientific uncertainty about (or even scientific support for) racial hierarchy. (Indeed, the origins of the "tobacco strategy" are often dated to a 1954 advertising campaign by the tobacco industry, 17 years after Pioneer was founded. Scientific racists and the eugenics movement might have been deploying the "tobacco strategy" for decades before the tobacco industry.)

The Pioneer Fund itself clearly exemplifies the strategies of supporting friendly research; and figures such as Shockley and Jensen exemplify the strategy of recruiting high-profile, distinguished scientists [@TuckerFundingScientificRacism2002].

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "bridge."  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be "incubated," protected from the destructive critical scrutiny they would receive in the mainstream. Bridge venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research.

MQ does not appear to have ever been considered a mainstream journal; it has always existed firmly on the margins. If MQ provided the "safe space" for "incubating" race science, away from mainstream criticism, then bridge venues would be necessary to disseminate race science into the scientific mainstream. We therefore hypothesized (H1) that scientometric and text mining analyses would show race science discourse appearing first and most prominently in MQ, and appearing in mainstream publications only subsequently, with certain venues serving as the bridge.

Specifically, among both (other) academics and the general public, the field of behavior genetics is strongly associated with race science [@PanofskyMisbehavingScienceControversy2014]. We therefore hypothesized (H2) that the journal *Behavior Genetics* (BG) would be a major bridge venue for the dissemination of race science discourse.

@@

After document retrieval and text extraction, we pre-processed the text using the `spaCy` NLP (natural language processing) Python library [@HonnibalSpaCyIndustrialstrengthNatural2018 version 1.9.0] and the R API `spacyr` [@BenoitSpacyrWrapperSpaCy2020 version 1.2.1].  Specifically, we applied regular expressions to remove header/footer copyright notices and hyphenation, used spaCy to annotate and extract noun phrases (eg, "the intelligence test items"), and then cleaned and standardized these phrases (eg, removing the/an/a, coverting all text to lowercase, and replacing all whitespace with underscores: "intelligence_test_items").  We then counted the occurrence of each noun phrase in the document.  The aggregated "document-term matrix" was stored in Parquet format [@VohraApacheParquet2016] for performance reasons, and written and read using the `arrow` package for R [@RichardsonArrowIntegrationApache2023 version 7.0.0].

@@

With respect to the aim of detecting and studying the prevalence of race science, a critical limitation of our methods is that they cannot distinguish race science from the broader category of race science discourse. Topic modeling represents a document as an unstructured "bag of words"; while it can detect that the document discusses both intelligence and racial groups, for example, it cannot detect whether the author endorses or critiques race-and-intelligence research. For the same reason, topic modeling cannot determine whether an article purports — implicitly or explicitly — to justify racial inequality and colonialism by appealing to scientific (or pseudo-scientific) findings, and thus it cannot distinguish race science discourse from scientific racism.  We therefore can only make claims about finding race science discourse, except when manual inspection or prior research indicates otherwise [e.g., @JacksonJr.ArthurJensenEvolutionary2022].

To generalize this limitation, the text mining techniques we use here are unable to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from overt and direct to a subtle "color-blind" racism [@Bonilla-SilvaRacismRacistsColorBlind2006]. Color-blind racism represents a change in the primary mechanisms by which white privilege is maintained, using non-racialized language in furtherance of the racial status quo both institutionally and individually, and increasingly reliant on the invisibility of socio-cultural mechanisms to preserve inequalities. Color-blind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" [@SaulDogwhistlesPoliticalManipulation2018].  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the color-blind language of "group differences" [@AlderferScienceNonsciencePsychologists2003].  These cases suggest further that color-blindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles [@GillbornSoftlySoftlyGenetics2016; @WillsAreClustersRaces2017].

@@

Recently, some computational scholars have argued that foundational large language models (LLMs) are able to encode broad cultural contexts [@UnderwoodWhyAINeeds2024] and produce sophisticated analyses of arguments [@CaulfieldTwoMeaningsReasoning2024]. If they are right, future LLM-based methods might possibly be able to distinguish supporters from critics of race-and-intelligence research, detect the use of racial dogwhistles, and perhaps even reliably identify the fuzzy boundaries of essentially contestable categories such as race science. However, such methods are beyond the scope of the project we present here.

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Topics 07 and 24 are examples of the two distinct types of race science discourse, while topic 22 is "mainstream" or non-race science intelligence research.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic; text version in table @tbl-silge.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal}

@@

In almost all years, most papers in topic 24 were published in mainstream journals rather than MQ.  Jensen's "How Much Can We Boost IQ and Scholastic Achievement?" [@JensenHowMuchCan1969] was published in 1969 in *Harvard Educational Review* (not included in this study), and the time series indicates that, in the early 1970s, there was an increase in articles in topic 24 in both PR and MQ (the only two journals in our corpus that were active at the time).  MQ shows another sharp increase in the late 1980s; using content analysis, Adams and Pilloud found that psychology was the dominant discipline in MQ in the period 1992-2018 [@AdamsMisappropriationBiologicalAnthropology2021].  PID published multiple papers in topic 24 almost immediately after it was founded, with Int showing a more gradual increase between the mid-1970s and mid-1990s.

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct version of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as venue for late 20th century race science discourse than mainstream psychology journals, especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].

## *[implications for agnotology]*

In terms of the margins-mainstream-bridge model of the "echo chamber effect," there was no particular bridge that carried ideas originating in MQ on the margins to mainstream journals. Within the scope of our data, the race-and-intelligence version of race science discourse was always a mainstream phenomenon.

The circulation of ideas is a prominent theme in the empirical agnotology literature.  Researchers have often examined the public circulation of ideas — how an industry attempts to manage public concerns about its products in marketing, press releases, and interviews — as well as the circulation of ideas within a company or industry itself — as in memos among intramural scientists and management [@ProctorGoldenHolocaustOrigins2012].  In a series of articles, Supran and Oreskes have shown that Exxon/ExxonMobil lied — knowingly made false statements — about climate change, contrasting skeptical and denialist statements in public-facing documents with skillful climate projections by the company's own scientists [@SupranAssessingExxonMobilsClimate2017; @SupranAddendumAssessingExxonMobils2020; @SupranRhetoricFrameAnalysis2021; @SupranAssessingExxonMobilsGlobal2023].

Less empirical research has examined the circulation of ideas within the scientific community.  @SupranAssessingExxonMobilsGlobal2023 include several peer-reviewed journal articles in their content analysis, but compare them to intramural documents and public statements, rather than the broader scientific literature.  @FernandezPintoKnowBetterNot2017 gives three examples to illustrate the "tobacco strategy": the joint public relations-research funding campaign by the tobacco industry itself; medical ghostwriting by the pharmaceutical industry; and efforts by conservative think tanks to promote a simplistic account of the causes of the Great Recession that shifted blame away from deregulation of the finance industry.  Both the first and last cases emphasize the circulation of ideas to the public, and the third does not involve academic researchers in any significant way.  Medical ghostwriting is concerning as a strategy for circulating false or misleading claims in the scientific literature. But it has not been a major case in the agnotology literature in recent years. This might reflect less concern about ghosting in biomedical research over time: on 2025-01-16, a Web of Science search for "medic* ghostwrit*" returned 85 total papers published between 2003-2024, with the number of publications per year peaking at 11 in both 2010 and 2011 and declining to 1-2 per year in 2023 and 2024.

Some researchers have used agent-based models (computer simulation methods) to study the circulation of misinformation within scientific communities under agnogenic ("ignorance-creating") influences such as "industry selection" and increased dissemination of favorable views [@HolmanExperimentationIndustrialSelection2017; @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. While these models can give insight into how certain mechanisms might work, it is difficult to extrapolate from them to actual cases [@FernandezPintoEpistemicLandscapesReloaded2018; @HicksOpenScienceReplication2023].

Understanding the public circulation of race science — its translation into scientific racism — is essential. But our findings suggest the circulation of race science within the scientific community warrants further investigation.  Using the set of race science discourse articles identified in this project as a starting point, future research might examine authorship and/or citation patterns.  For example, have race science discourse articles primarily produced by a small and stable set of authors, or a larger set that has changed over time?  Or, to what extent have race science discourse articles been cited within non-race science articles?

## *[implications for psychology]*

@@

Over the past few decades, many professional organizations in genetics and anthropology have made formal statements rejecting race as a biologically meaningful concept [@anthropology1996aapa; @american2018ashg; @board1998aaa; @committee2020asa; @FuentesAAPAStatementRace2019; @nhgri2023; @rotimi2022ashg; @wynshawboris2020ashg], undermining the core assumption of race science.  The American Psychological Association (APA) --- whose members are predominantly clinical psychologists --- the Federation of Associations in Behavioral & Brain Sciences (FABBS) --- an organization whose purpose is to provide policy recommendations --- and the editors of *Nature Human Behavior* have openly rejected race as biologically meaningful [@american2021apology; @baron2022fabbs; @EditorsScienceMustRespect2022].  But research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as biologically meaningful when they denounce racism [@aps2021statement; @board2020psychonomic].  Psychology as a discipline maintains space for several mythological race science narratives, such as allegations of a taboo against race and intelligence research [@jackson2021mythical], claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science [@roberts_2022], or arguments that holding race science to the same evidentiary standards as other psychological research is a violation of academic freedom [@herbert2023academic].  And historically several race scientists have secured important gatekeeping positions within the scientific community, such as on the editorial boards of *Intelligence* and *Personality and Individual Differences*.

@@

# Operationalizing "race science" in practice {#sec-operationalizing}

A reviewer raised a concern that our conception of race science involved a counterfactual that "could be difficult to confirm or refute." A full discussion of how, for example, a research funder might operationalize race science in order to subject it to heightened critical scrutiny before funding, is beyond the scope of the current project. Here we sketch three frameworks that might be used to operationalize race science, either independently or in combination.

First, some research has actually been utilized in actual scientific racist propaganda, for example, in memes that circulate on fringe discussion boards or as citations in the manifestos of mass shooters. This research is ipso facto race science. Taken on its own, this framework is only useful after the fact. But it might still play a substantial role in, for example, a decision whether to retract a publication or adjudicate whether a researcher has engaged in misconduct.

This first framework can also provide important evidence for the second framework, which draws on the *reasonable person* standard from the law of torts. In other words, the second framework asks whether a reasonable scientific peer — well-informed about what kinds of research have been utilized by scientific racists and how, and concerned to avoid this happening in the future — would consider the research likely to be utilized in a similar way in the future. @DouglasSciencePolicyValuefree2009 argues that, because we all have a moral responsibility to exercise reasonable foresight over our activities, scientists have a specific responsibility to exercise such foresight over the social consequences of their research. Reasonable foresight does require projecting counterfactual scenarios; but these can be based in casuistical reasoning (comparisons with actual cases, as per the first framework) and are not different in kind from the projections made by research funders and hiring committees.

@PenceChallengesCommunityScience2023 argue that a reasonable foresight framework encounters difficulties as scientific communities become more nebulous and vaguely defined, and propose that exercising reasonable foresight should include community engagement and incorporating outsider perspectives. So, specifically, cognitive psychologists might make a point of engaging with groups that have been marginalized by race-and-intelligence research and working with scholars who have studied how and why certain kinds of research are taken up by scientific racists.

The third framework starts with exactly this question of how and why research is taken up by scientific racists. In other words, what features of a project make it race science?

Schematically, a research project can be thought of as using *methods* to produce *findings* that (aim to) answer *research questions*. Each of these three elements requires various assumptions to be logically relevant to the other two  [@LonginoScienceSocialKnowledge1990; @LloydAdaptationismLogicResearch2015]. For example, a particular project might use data from a national genome bank and a genome-wide association study (GWAS; method) to develop a polygenic score estimator of educational attainment (finding) with the aim of answering the question "To what extent is cognitive ability genetic?" This project must assume that a sample from a single nation is representative of humanity in general (to be relevant to the generalized framing of the question), that "cognitive ability" is a well-defined construct and worth investigating, that educational attainment is an accurate measure of cognitive ability, that the SNPs included in the genome data capture genetically-caused variation, that analysis of the polygenic scores can recover/measure this variation, that variation can be partitioned into genetic and environmental causes [@KellerMirageSpaceNature2010], and so on.

Some of these assumptions — or the findings themselves — can support scientific racist claims, either through strict deductive entailment (the kind studied in formal logic courses) or through *implicature* [@DavisImplicature2024]. The concept of "implicature" refers to the phenomenon that language use goes beyond the strict deductive entailments — "we mean more than we say." Consider the following exchange between a teenager and their parent.

| 
| Teen: Sasha's going to Cold Stone. 
| Parent: The dishes are still in the sink. 
| Teen: I don't have practice in the morning. 
| Parent: Be home by 10.

Taken strictly, the exchange is nonsensical: Sasha going to Cold Stone has nothing to do with the dishes, which in turn has nothing to do with whether the teen has practice or what time they should be home.

Instead the exchange runs on implicature. The teen's first statement implies but does not strictly entail a request, that they be granted permission to go to Cold Stone as well. The parent's first response is an implied denial of this request, on the grounds that the teen hasn't done their chores. The teen responds with an implied promise to finish their chores in the morning, and the parent implies permission with a proviso.

Implicature is context-dependent; the exchange above assumes that the teen wants to do the same thing as Sasha, that they're responsible for doing the dishes, usually have practice in the mornings, and so on.  In a different context, without this assumptions in play, the implicature doesn't work.

We can think of a social context as providing unstated assumptions that serve as additional premises, and thereby allow language users to draw additional inferences from a statement that go beyond its strict entailments. Returning to the hypothetical GWAS of educational attainment study, suppose the study found that "educational attainment is 21% heritable." This supports the simplification "educational attainment is substantially heritable." Then, in a context in which educational attainment is assumed to reflect intelligence, this has the implicature "intelligence is substantially heritable." Next, with the genetic/environmental partition assumption and in a context where genetic determinism is widely held, this has the further implicature that "intelligence is innate." Combined with studies that claim differences in mean IQ across racial groups (or even just a contextual assumption to this effect), an assumption that IQ also measures intelligence, and statistical essentialism (the idea that statistical averages reflect group essences), we get to the scientific racist implicature that "some races are innately more intelligent than others."

In this way, a study that did not mention race — or, indeed, might have mentioned race only to highlight that the sample was racially homogeneous and therefore the results could not be generalized — could still be utilized to support scientific racist implicatures and therefore be race science [@KitcherArgumentFreeInquiry1997; @WillsAreClustersRaces2017; @BirdTypologicalThinkingHuman2024]. (Note that denying the generalizability of results entails that the study is not deductively relevant to the generalized research question.)

Our example in the main text, of the Flynn effect as legitimizing race science even though it poses an empirical challenge to certain aspects of race science (hereditarianism), can be analyzed in a similar way. A project investigating the Flynn effect might pose a research question that assumes — implicitly or explicitly — that it is scientifically possible but not certain than hereditarianism is false. This entails that it is scientifically possible but not certain that hereditarianism is true. Then, in a context where it is widely assumed that any scientifically possible but uncertain claim is a legitimate subject for scientific research, this supports the implicature that hereditarianism is a legitimate subject for scientific research. In this way, Flynn effect research can be race science discourse.

Because key assumptions that enable scientific racist implicatures are often implicit and context-specific, it will often be difficult to empirically prove to the satisfaction of everyone concerned that a particular project is (or is not) race science — outside of cases where the research has actually been utilized in actual scientific racist propaganda. Authorities within scientific communities — funders, journal editors, hiring and tenure committees — will need to decide what kinds of evidence are appropriate for adjudicating claims of race science, and whether there is sufficient evidence to reach a conclusion (one way or the other) in any given case. The frameworks we have offered here can help these authorities systematically and empirically — that is to say, scientifically — adjudicate such cases.  But race science is still an unavoidably political concept.
